Bayesian Inference from Time Series: Information Loss and the Whittle Likelihood

### 2018-03-05 doc/ms.tex
PSD
More writing

@@

We assume that the probability distribution for $\epsilon(t)$ is indeed white
noise. In this case, the probability distribution for $y(t)$ given the
parameters is a gaussian process with mean function of zero and
covariance function given by
\begin{align}
C(\tau) &= A^2\left[\cos(\eta\omega_0\tau)
                + \frac{1}{2\eta Q}\sin(\eta\omega_0\tau)\right]
\end{align}
where $\eta = \sqrt{1 - 1/(4Q^2)}$, $\tau$ is a time separation,
and $Q > 1/2$ (i.e., we assume the oscillation is not overdamped).
A more general expression which allows for $Q \leq 1/2$
is given by \citet{anderson} and \citet{celerite}.
In the limit of large $Q$, the covariance function is approximately
the product of an oscillation and an exponential decay:
\begin{align}
C(\tau) &= A\exp\left(-\frac{\omega_0\tau}{2Q}\right)\cos(\omega_0\tau)
\end{align}
which was the form assumed by \citep{brewer_stello}.

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
20          &       Exact           & \\
20          &       Whittle         & $1.491 \pm 0.055$\\ 
50          &       Exact           & \\
50          &       Whittle         & $1.957 \pm 0.055$ \\
100         &       Exact           & $2.605 \pm 0.053$ \\
100         &       Whittle         & $2.052 \pm 0.055$ \\
200         &       Exact           &   \\
200         &       Whittle         &   \\
\bottomrule
\end{tabular}
\caption{\label{tab:oscillator_results}}
\end{table}

@@

@@

\bibitem[Foreman-Mackey et al.(2017)]{celerite}
Foreman-Mackey, D., Agol, E., Ambikasaran, S. and Angus, R., 2017. Fast and scalable Gaussian process modeling with applications to astronomical time series. The Astronomical Journal, 154(6), p.220.
\end{thebibliography}

>>>>

\section{Introduction}

In a recent paper, I introduced a computational technique based on Nested
Sampling for estimating the entropy of probability distributions which are
usually only accessible through sampling, such as posterior distributions.

@@

We assume that the probability distribution for $\epsilon(t)$ is indeed white
noise. In this case, the probability distribution for $y(t)$ given the
parameters is a gaussian process with mean function of zero and
covariance function given by
\begin{align}
C(\tau) &= A^2\left[\cos(\eta\omega_0\tau)
                + \frac{1}{2\eta Q}\sin(\eta\omega_0\tau)\right]
    \label{eqn:oscillation_covariance}
\end{align}
where $\eta = \sqrt{1 - 1/(4Q^2)}$, $\tau$ is a time separation,
and $Q > 1/2$ (i.e., we assume the oscillation is not overdamped).
A more general expression which allows for $Q \leq 1/2$
is given by \citet{anderson} and \citet{celerite}.
In the limit of large $Q$, the covariance function is approximately
the product of an oscillation and an exponential decay:
\begin{align}
C(\tau) &= A\exp\left(-\frac{\omega_0\tau}{2Q}\right)\cos(\omega_0\tau)
\end{align}
which was the form assumed by \citet{brewer_stello}.

The power spectral density corresponding to
Equation~\ref{eqn:oscillation_covariance} is
\begin{align}
S(\omega) &= \sqrt{\frac{2}{\pi}}
             \frac{S_0\omega_0^4}
                  {(\omega^2 - \omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{align}
where $S_0 = A/(\omega_0Q)$.

\subsection{Inferring the period}
The parameter $\omega_0$ is the angular frequency, related to the period
$T$ by
\begin{align}
T &= \frac{2\pi}{\omega_0}.
\end{align}

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
20          &       Exact           & \\
20          &       Whittle         & $1.491 \pm 0.055$\\ 
50          &       Exact           & \\
50          &       Whittle         & $1.957 \pm 0.055$ \\
100         &       Exact           & $2.605 \pm 0.053$ \\
100         &       Whittle         & $2.052 \pm 0.055$ \\
200         &       Exact           &   \\
200         &       Whittle         &   \\
\bottomrule
\end{tabular}
\caption{The mutual information between the data and the log-period, for
different numbers of datapoints $N$ and for different likelihoods (exact vs.
Whittle).
This is equal to the prior expected reduction in the entropy
in going from the prior to the posterior.\label{tab:oscillator_results}}
\end{table}

@@

\bibitem[Brewer(2017)]{brewer}
Brewer, Brendon J. ``Computing Entropies with Nested Sampling.'' Entropy 19, no. 8 (2017): 422.

@@

\bibitem[Foreman-Mackey et al.(2017)]{celerite}
Foreman-Mackey, Daniel, Eric Agol, Sivaram Ambikasaran, and Ruth Angus. ``Fast and scalable Gaussian process modeling with applications to astronomical time series.'' The Astronomical Journal 154, no. 6 (2017): 220.
\end{thebibliography}
